A Recipe for Scavenging - the natural history of a behaviour

### 2016-04-05 natural history of scavenging KH edits.tex
Included extra doc KH edit with lots of changes

\subsection*{The Difficulty of Scavenging} %KH: Make that the template for all the next sections
The chief hurdle to scavenging is finding a sufficient quantity of food, the occurence of which is difficult to predict in space and time. Thus, any animal existing as a scavenger must minimise its locomotory costs and maximise its detection capabilities \citep{ruxton2004obligate}.
Once found, the scavenger has to process the carrion and overcome the agents of decay produced by the action of microorganisms on the carcass in order to take in a suffienct amount of energy \citep{ruxton2014fruit}. 
The habitat must also be productive enough to sustain a pool of animal biomass that will eventually produce carcasses.
These facets are essentially the key parameters found in the species functional response curves of optimal foraging theory, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}. 
Here, we can draw on these aspects to envisage a scavenger moving through its environment, searching for food and trying to process it efficiently as we explore the prevalence of scavenging through time. 
Our aim is to produce a 'scale of scavenging' whereupon we can place any species, past or present, and assess the importance of scavenging in its diet. 
% The idea of scrounging from predator kills is undermined by studies showing that in the  majority of ecosystems more animals die from disease and starvation than predation \citep{benbow2015introduction}.

@@

%KH: need paragraph for how do we know something is (was) a scavenger
%TG: Insert figure here: scavenging scale + four points: Locomotion/Detection/Consumption/Environment
\section*{Encounter Rate}

\subsection*{Locomotion}
As endotherms, mammals can sustain long bouts %TG: not sure about "bouts" get it with the context and I know it makes the whole review a bit more literary and thus nicer to read but it might exclude non-native english readers. 
 of energetically expensive activity \citep{bennett1979endothermy}. 
By contrast, modern reptiles are ectothermic, limiting their activity periods \citep{bennett1979endothermy}.
% a successful reptilian scavenger requires a far different set of adapations.
This is exacerbated by the sprawling gait seen in many lizards which results in the phenomenon known as Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}.
These constraints manifest in aspects such as maximum sustainable speed where an equivalent mammal enjoys a six to seven fold increase \citep{ruben1995evolution}.
To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass.
In the case of mammals and reptiles these are 1.15 $\times$ body mass (kg) \textsuperscript{0.12} and 0.23 $\times$ body mass (kg) \textsuperscript{0.12} respectively \citep{ruxton2004obligate}.
If we insert these into a foraging radius model $\sfrac{\frac{\text{duration} \times \text{speed}}{2}}{1000}$ for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km, an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}. 
For a foraging scavenger, this ability translates into a greater area searched for food.

@@

Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs.
They were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
Of relevance, are the questions that still persist about their metabolism, with the latest evidence suggesting they were mesothermic i.e. intermediate to ecto- and endotherms \citep{grady2014evidence}. 
We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}. %TG: and cite "Kane et al in press" no? I think these people did some kind of work on that matter.
Taken together, this implies dinosaurs had a foraging range that fell in between that of modern terrestrial mammals and reptiles.

Of course, the importance of terrestrial tetrapods predates the evolution of the dinosaurs. %TG: I will down that "dominance" term. Maybe change the sentence to "Of course the importance of terrestrial tetrapods in ecosystems predates the evolution of the dinosaurs." AK Done 
It is during the Permian (298.9 - 252.17 Mya) that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
And it is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion.

Scavenging behaviour may have evolved on land as soon as the first terrestrial tetrapods emerged.
In fact, some of the earlier tetrapods tracks dating back to the early Middle Devonian (393.3 - 387.7 Mya) were found in intertidal environments \citep{Niedzwiedzki2009}.
These environments are isolated from marine systems twice a day leaving potential carrion unexploited by marine vertebrates.
\cite{Niedzwiedzki2009} suggest that these environments ``would thus have allowed marine ancestors of tetrapods gradually to acquire terrestrial competence while accessing a new and essentially untouched resource.''

%TG: I think it's missing a transition here: maybe something along "But moving on land as it's limitations..."
But it is in the air that we find scavengers \textit{par excellence}. %TG: not sure how reviewers will go with French but I like it!
Certainly, scavenging should be particularly attractive to avian predators compared to mammals. % Not sure about this paragraph 
The latter can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier \citep[e.g. socially hunting lions; ][]{owen2008predator}.
In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey} because of the greater cost of injury and need to carry off their food.
% This is likely due to their need to kill animals that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal.
Scavenging provides a means for birds to exploit species that would otherwise be too big for them to kill.
% Thus, for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.

Flight is a cheaper means of locomotion than either walking or running \citep{tucker1975energetic}. 
We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}.
The advantage of flight can be extended further in larger species that engage in soaring instead of flapping flight, which is even cheaper still for searching the environment \citep[approximately twice the basal metabolic rate; ][]{hedenstrom1993migration,spivey2014analysing}.
The benefits this confers are clear from the information we have on the enormous foraging ranges of many vultures \citep{spiegel2013factors} and seabirds \citep{thaxter2012seabird}.
The difficulty inherent in observing marine birds foraging at sea means our knowledge of their feeding habits is limited. 
But it does seem that albatrosses, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}.
This is typically in the form of squid, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}. 
In vultures, who are more easily observed on land, we have the best known scavengers on Earth. 
These birds consist of two convergent groups, from the old and the new world where they represent the only example of obligate vertebrate scavengers to date. %TG: Or "obligate vertebrate scavengers known to date." For a more emphasizing turn.

%TG: I thought the paragraph was a bit long so here's a cut for "flying scavengers in the past"
The two families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Mya; ][]{Jetz2012, Jarvis2014}.
Yet, avian flight is far older than this and originates in the Late Jurassic (163.5-145 Mya), conincident with the fossils of \textit{Archaeopteryx lithographica}, so many of the benefits of flight would have been realised from that point on for carnivorous birds.
Vertebrate flight is much older still, where pterosaurs predate bird origins by a considerable margin in the Late Triassic (235-201.3 Mya).
Scavenging in this diverse group has been hypothesied many times before \citep{witton2008reappraisal}.
Certain clades of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.
%TG: this paragraph's a killer! Love it!

The only other vertebrate group capable of powered flight are the bats where scavenging has not been recorded to our knowledge.
However, some fossil from the Eocene (56 to 33.9 Mya) suggest the presence of possible scavenging bats.
A large bat from the middle to late Eocene (56 to 33.9 Mya) with a robust cranio-mandibular morphology, \textit{Necromatis} (``death-eater''), could have intake carrion as a significant proportion of its diet \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}.
Although, this example might be peculiar to this exceptional fossil given most other remains represent animals less suited to the lifestyle \citep{eiting2009global}.

Similar to aerial species, aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. 
In fact, the cost of swimming is lower than either running or flying \citep{williams1999evolution}. 
This has led some researchers to argue for the feasibility of an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. 
As in aerial and terrestrial environments we have evidence of facultative scavenging among extinct aquatic species.
For example, shark teeth have been found in the remains of dinosaurs, mososaurs and 4-million-year-old mysticete whales \citep{schwimmer1997scavenging,ehret2009caught}.  
%the remains of a mosasaur and a terrestrial hadrosaur were discovered with embedded teeth from a Cretaceous shark, \textit{Squalicorax} \citep{schwimmer1997scavenging}.
%As well as a likely instance of scavenging between a 4-million-year-old white shark (\textit{Carcharodon}) and  whale from Peru \citep{ehret2009caught}.
Extant White sharks (\textit{Carcharodon carcharias}) too are known to feed on whale carcasses \citep{fallows2013white}.

We might expect then that by combining an aquatic environment and an endothermic metabolism that marine mammals would prosper as scavengers. 
Fossil pinnipeds and cetaceans from 60 Mya have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, their total energetic cost is similar to an equivalent terrestrial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an expensive system.

@@

>>>>

%the intro needs to lead towards a aim a bit more strongly

\subsection*{The Difficulty of Scavenging} %KH: Make that the template for all the next sections

The chief hurdle of scavenging is finding a resource that is often difficult to predict in space and time. Though chance alone many species will avail of some level of opportunistic scavenging opportunities. However, species that rely on scavenging to sustain substantial portions of their diets must increase the probability of encountering a sufficient amount of carrion in order to meet some energetic demands. Once found, the scavenger must be able to out-compete any potential competitors and process a product under the processes of decay from toxin producing and disease causing microorganisms \citep{ruxton2014fruit}. Finally, the level of scavenging opportunities available will also depend on the density, size, and quality of carcases produced, all of which are dependent on complex ecosystem dynamics. These facets which are required for scavenging are essentially the key parameters found in the species functional response curves of optimal foraging theory, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}. We compare the ability of scavenging across the vertabrates based on each of these facets and their relationship with scavenging. By framing scavenging in a framework of optimal foraging and identify the perquisite attributes and processes required for scavenging we produce a 'scale of scavenging' whereupon we can place any species, past or present, and assess the importance of scavenging in its diet.

@@

%KH: need paragraph for how do we know something is (was) a scavenger
%TG: Insert figure here: scavenging scale + four points: Locomotion/Detection/Consumption/Environment
\section*{Encounter Rate}
All foraging processes depend on encounter rates between the consumer and the resource. In the simplest case, encounter rates can be thought of in terms of gas diffusion models were the movement of two agents depends on only on the relative speed of the agents. As carcasses are prone to being relatively stationary, the relative speed between a scavenger and carrion is dependent on the movement of the scavenger. As most vertebrates depended on more than simply bumping into resources like gas molecules and can actively detect resources through sensory abilities, scavenging will also depend on search rates which are dependent on both species physiology and the dimensionality of the environment (You guessed it Pawar et al 2012).

\subsection*{Locomotion}
In contrast to the importance of locomotion in predation, locomotion relating to scavenging depends more on the ability to efficiency move over large areas. This ability generally requires an efficient transfer of metabolic energy into movement which may rely on both physiology (i.e. metabolism) and the medium of the environment in which they are moving (aerial, aquatic or terrestrial).

Perhaps the most efficient form of locomotion in vertebrates is paradoxically found in flying species. Despite the energetic costs of flight the only known vertebrate obligate scavengers are the old and the new world vultures. While powered flight is energetically expensive (ref) it is the ability of these species to exploit air currents using large wingspans that allows them to soar while only approximately doubling their metabolic rate \citep{hedenstrom1993migration,spivey2014analysing}. By depending on thermal air flows these species can forage over enormous ranges \citep{spiegel2013factors}. This mode of locomotion is also exploited by sea birds, which use strong ocean winds to search large areas of the oceans \citep{thaxter2012seabird}. While many species of seabird are likely primarily predators, it seems that albatross, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}. This is typically in the form of squid, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}.

The use of soaring in extinct species as a means for scavenging is also likely. The two families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Mya; ][]{Jetz2012, Jarvis2014} and are  likley to have being soaring flyer's as demonstrated by their large wingspans (Is this true?). (Is there any other form of soaring?). However, soaring flight is likely to be far older than these groups with avian flight originating in the Late Jurassic (163.5-145 Mya) and pterosaurs in the Late Triassic (235-201.3 Mya). Scavenging in the diverse pterosaurs group has been hypothesied many times before \citep{witton2008reappraisal}. Certain clades of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.

While soaring flight is perhaps the only viable means of locomotion that allows for an obligate scavenging life-style (ref Ruxton) powered flight is still an efficient means of locomotion. Flight is a cheaper means of locomotion than either walking or running \citep{tucker1975energetic} (This needs to be much clearer, needs units etc). We know that many extant birds exist as facultative scavengers because storks, raptors and corvids all take substantial quantities of carrion in their diet \citep{kendall2013alternative}. Similarly we would expect that extinct species would also gain in a similar fashion depending on the efficiency of flight. For example, early flying species such as archaeopteryx are predicted to be relatively inefficient (Is there anything on this as I imagine early flight was diverse). Similarly someithing abou t the effecieny of pterosuars flight. The importance of efficient flying over large areas may explain the lack of scavenging behavior in bats as they are general smaller and less efficient in comparison to larger birds (speculating here but I wouldnt be surprised), with only the textit{Necromatis} (``death-eater''), a large bat from the middle to late Eocene (56 to 33.9 Mya) with a robust cranio-mandibular morphology, a likely candidate for scavenging behavior amongst the x amount of known bat species \citep{Weithofer_Necromantis_1887,Hand_Necromantis_2012}.

Similar to aerial species, aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. In fact, the cost of swimming is lower than either running or flying \citep{williams1999evolution} with only soaring flight likely to surpass it (I assume the reference is for powered flight, again this needs to be more precise). This has led some researchers to argue for the feasibility of an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. Sharks perhaps best resemble the large soaring flyer's as they depend on large pectoral fins in order to maintain lift. Shark species show very large ranges (for example great white sharks) and are amongst the most lowest cost swimmers (I know the Greenland shark is the slowest swimmer). Faculative scavenging is seen in many marine groups including species of extant sharks such as White sharks (\textit{Carcharodon carcharias}) feeding on whale carcasses \citep{fallows2013white}, Greenland sharks feeding on seals (refs), sixgill sharks (the ref with the pigs) and also extinct species with shark teeth found in the remains of dinosaurs, mososaurs and 4-million-year-old mysticete whales \citep{schwimmer1997scavenging,ehret2009caught}.

Other types of swimming in fish, other species propelled by fins (turtles and plesiosaurs etc)

snakes - ie undulating

endotermy - why it doesn't seem to work
We might expect then that by combining an aquatic environment and an endothermic metabolism that marine mammals would prosper as scavengers.
Fossil pinnipeds and cetaceans from 60 Mya have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, their total energetic cost is similar to an equivalent terrestrial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an expensive system.

Terrestrial environments are the most energetically costly in which to move, as a animal must trade off the cost of holding the body up and propelling itself forward. Species which do not support an upright position such as modern reptiles limit their ability to move for sustained periods \citep{bennett1979endothermy}. This is exacerbated by the sprawling gait seen in many lizards which results in the phenomenon known as Carrier's Constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). This would also be true of extinct species. It is during the Permian (298.9 - 252.17 Mya) that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
Snakes are a bit different are more effecent at moving I think - laots of scavanging. Both these groups are more lilky to depend on low energetic costs in order to componsate for low ecnounter rates duee to thier inabilty to move. It is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion.

Modern endothermic mammals can sustain longer periods of energetically expensive activity \citep{bennett1979endothermy} resulting in larger foraging ranges. To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass.
In the case of mammals and reptiles these are 1.15 $\times$ body mass (kg) \textsuperscript{0.12} and 0.23 $\times$ body mass (kg) \textsuperscript{0.12} respectively \citep{ruxton2004obligate}.
If we insert these into a foraging radius model $\sfrac{\frac{\text{duration} \times \text{speed}}{2}}{1000}$ for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km, an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}. 
For a foraging scavenger, this ability translates into a greater area searched for food.

@@

Something more a bipedal movement.
Unsurprisingly, given their enduring appeal, the prevalence of scavenging has been explored in the carnivorous, theropod dinosaurs.
They were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
They are quite alien to anything we know today which restricts our ability to understand their ecology far more so than extinct mammals \citep{weishampel2004dinosauria}.
Of relevance, are the questions that still persist about their metabolism, with the latest evidence suggesting they were mesothermic i.e. intermediate to ecto- and endotherms \citep{grady2014evidence}. 
We do know that they walked with the erect gait of mammals or birds rather than the sprawling gait of lizards and that they were most likely facultative scavengers \citep{weishampel2004dinosauria,depalma2013physical}. %TG: and cite "Kane et al in press" no? I think these people did some kind of work on that matter.
Taken together, this implies dinosaurs had a foraging range that fell in between that of modern terrestrial mammals and reptiles.

Something about humans (also the carnivours kangaroo which might be the most efficient mover.)

@@

%%%%%%%%%%this is moved from the locomotion section as it didn't seem to fit.
Certainly, scavenging should be particularly attractive to avian predators compared to mammals. % Not sure about this paragraph 
The latter can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier \citep[e.g. socially hunting lions; ][]{owen2008predator}.
In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey} because of the greater cost of injury and need to carry off their food.
% This is likely due to their need to kill animals that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal.
Scavenging provides a means for birds to exploit species that would otherwise be too big for them to kill.
% Thus, for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.
%%%%%%%%%%this is moved from the locomotion section as it didn't seem to fit.
